Bayesian model in JAGS

### 2023-01-20 3_JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
add new

>>>>

## A generalized linear mixed effect model (GLMM) framework

To produce accurate estimates of disease incidence, statistical models must take into account the full range of features present in the data, including non-normal distributions of count data, the impact of explanatory variables, and spatial correlation. One solution is to utilize generalized linear mixed models (GLMMs), which integrate a generalized linear model (GLM) with normally distributed random effects that account for both spatial correlation and overdispersion.
\
\
An example model structure is given as:
$$Y_i \sim Poisson (\mu_i)$$
$$\log \mu_i= \log E_i+x_i^\prime \alpha+\phi_i$$
where $Y_i$ is the observed number of disease in region $i$, $\mu_i$ is the  conditional mean, $E_i$ is the expected number of disease in region $i$, $x_i$ is the explanatory variables (covariates) of region $i$, $\alpha$ is the coeffiecients of the corresponding explanatory variables, and $\phi_i$ accounts for the random effects, which can be spatially correlated.

## Conditional autoregressive (CAR) priors for spatial modeling

To model spatial autocorrelation, four CAR priors are commonly used, including the intrinsic models, BYM models, Leroux models, and Cressie models. In [**Duncan Lee's paper**](https://www.sciencedirect.com/science/article/pii/S1877584511000049?via%3Dihub), he systematically compared these four different priors and concluded that Leroux model is the best from both theoretical and practical standpoints. For short, Leroux model is appealing because it uses a single set of random effects to model the varying strengths of spatial autocorrelation $\rho$, where $\rho=0$ means spatially independent while $\rho$ near 1 indicates strong spatial correlation.

## Leroux priors for spatial modeling

The Leroux version of spatial random effects is given as:
$$\phi_i | \boldsymbol{\phi_{-i}},\rho,\tau^2 \sim MVN (\frac{\rho\sum_{j=1}^nw_{ij}\phi_i}{\rho\sum_{j=1}^nw_{ij}+1-\rho},\frac{\tau^2}{\rho\sum_{j=1}^nw_{ij}+1-\rho})$$
where $\rho \in [0,1)$ represents the spatial correlation that was estimated from the data; $\rho=0$ corresponds to spatial independence, simplifying the model to an independent random effects model, and $\rho$ near 1 indicates strong spatial correlation. The $w_{ij}$ are variables that contain information about spatial proximity between two locations; $w_{ij}$ equals one if regions i and j share a common border and equal zero otherwise ($w_{ii}$=0 for all i). The total variance of spatial random effects is described by $\tau^2$.

## Create a spatial adjacency matrix 
###  Step 1. Set up -- Shapefiles

(I learned this from Kayoko Shioda!)

```{r,message=FALSE,warning=FALSE}
# Load a shapefile. Make sure each polygon on shapefile  
# corresponds to only one state. 
require(rgdal) # for working with shapefiles
require(ggplot2) # for plot and fortify
CT1map = readOGR(dsn=path.expand('./WGS84'), layer= 'zipct_37800_0000_2010_s100_census_1_shp_wgs84')
CTmap.df <- as.data.frame(CT1map)
CT1map_fort <- fortify(CT1map,region="ZCTA5CE10")
# ZCTA5CE10 is CT zipcode ID
# The ID variable is then the zipcode/area unit 
CT1map_fort$zipcode <- CT1map_fort$id

# Sort both the shapefile and the data file by ID.
# This is the most important step
CTmap.sort <- CT1map[order(CT1map$ZCTA5CE10),]
```

## How to create a spatial adjacency matrix

###  Step 2. Create a neighbor matrix from the sorted shapefile 
```{r,message=FALSE}
library(spdep) # for neighbor evaluation

# Neighbors can either be Queen 
# (any zip that touches another zip - even at corners) 
# or Rook neighbors (share full sides -- not corners)

# The snap dist is governing the NA in neighbor matrix
# Can remove SNAP statement; 
# if Queen = F then single point touch not allowed 
# and Rook distance used.

# When data is missing -- there's more non-zero links involved 
# Try to make snap distance as small as possible 
# while making sure all zipcode have at least one neighbor
neighb <- poly2nb(CTmap.sort, queen = T, snap = sqrt(0.001))

# Check average number of links -- increase snap distance 
# if it says there is a zip with zero neighbors
neighb # Avg Number of links = 8.18705

# Make neighbor matrix 
# if zero.policy FALSE stop with error for any empty neighbour sets, 
# if TRUE permit the weights list to be formed with zero-length weights vectors
# B is the basic binary coding, W is row standardised (both are commonly used)
neighb.mat <- nb2mat(neighb, zero.policy = T, style = 'B')
```

## How to create a spatial adjacency matrix

###  Step 3. Create a W matrix for Leroux model 
```{r,message=FALSE}
# For Leroux model, neighbor matrix needs to have 0 
# if not neighboring, -1 if neighbor, 
# and number of neighbors for each zip on diagonal), so...
w.mat <- ifelse(neighb.mat==1,-1,neighb.mat)
num.neigh <- colSums(neighb.mat) 
# Count the number of neighbors for each state
for (i in 1:nrow(w.mat)) {
  # Put the number of neighbors on diagonal
  for (j in 1:ncol(w.mat)) {
    if (i==j) {
      w.mat[i,j] <- num.neigh[i]
    }
  }
}
```
